Fire danger: the skill provided by the ECMWF Integrated Forecasting System

### 2019-11-21 article/article.Rmd
Added figures1-3
Updated the article with new method to retrieve forecasts

```
\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{station_error.png}
    \caption{Comparison between modelled FWI and observed FWI value. FWI are calculated from using ECMWF reanalysis (ERA5) and forecasts at different lead times. The box plots are used to describe the distribution of values accross the observation points for the 2017 Mean Bias  (panel a) and the 2017 mean absolute error (panel b)} \label{fig:synop}
\end{figure*}
```

Despite its importance  the analysis performed using the synop network it is still pointwise and does not cover all the regions where fires are relevant. Moreover mean biases and MAE are based on the high resolution forecast. These skill metrics do not provide information about the performance of the ensemble forecasting as a whole.

Figure \ref{fig:crp}

\textbf{ PUT here crps skills scores}

\subsection{ Skill in detecting fire events}

>>>>

```{r prepFigure3, echo = FALSE, message = FALSE, results = 'hide', cache = TRUE}
library("easyVerification")
# The following code uses forecast data which is restricted.
# A summary table is computed and saved for users at the end of this code chunk.
# Get data from previous publication
# https://github.com/cvitolo/GEFF-ERA5/blob/master/data/df_geff_erai_era5.rds
# save it in the "../data" folder
df <- readRDS(file = "../data/df_geff_erai_era5.rds")
# insert 1-year fire season for tropics (lat = [-30, +30])
df$season[df$lat < 30 & df$lat > -30] <- "Dry"
# Remove wet season (we are only interested in the dry season)
df <- df[df$season == "Dry", ]
# How many stations remain?
stations <- unique(df[, c("id", "lat", "long")])

df$region <- sapply(strsplit(df$tzid, "/"), `[`, 1)
df$date <- as.Date(paste0(df$yr, "-", df$mon, "-", df$day))
df <- df[, c("id", "lat", "long", "region", "date", "OBS", "ERA5")]

dates2017 <- seq.Date(from = as.Date("2017-01-01"),
                      to = as.Date("2017-12-31"),
                      by = "day")

forecasts_folder <- "/hugetmp/FireEventsDatabase/forecasts/fwi/"

df_used <- data.frame(matrix(NA, nrow = 0, ncol = 13))
names(df_used) <- c("id", "lat", "long", "region", "date", "OBS", "ERA5",
                    "HRES_d2", "HRES_d6", "HRES_d10", "CRPS_d2", "CRPS_d6", "CRPS_d10")

for (i in seq_along(dates2017)[216:365]){

  print(dates2017[i])

  # What stations have data on this date?
  dfx <- df %>% filter(date == dates2017[i]) %>% filter(complete.cases(.))
  # Define spatial point
  spdf <- dfx[, c("id", "lat", "long")]
  # Promote data.frame to spatial
  coordinates(spdf) = ~long + lat

  # Calculate lead days
  day2 <- dates2017[i] - 1
  day6 <- dates2017[i] - 5
  day10 <- dates2017[i] - 9

  # Get data from HRES forecasts, e.g. ECMWF_FWI_20170101_1200_hr_fwi.nc
  if (day10 %in% dates2017){

    # Day2
    HRES2 <- raster::brick(file.path(forecasts_folder,
                                     paste0("ECMWF_FWI_",
                                            gsub("-", "",
                                                 as.character(day2)),
                                            "_1200_hr_fwi.nc")))[[2]]
    dfx$HRES_d2 <- raster::extract(x = HRES2, y = spdf)
    # Day6
    HRES6 <- raster::brick(file.path(forecasts_folder,
                                     paste0("ECMWF_FWI_",
                                            gsub("-", "",
                                                 as.character(day6)),
                                            "_1200_hr_fwi.nc")))[[6]]
    dfx$HRES_d6 <- raster::extract(x = HRES6, y = spdf)
    # Day10
    HRES10 <- raster::brick(file.path(forecasts_folder,
                                      paste0("ECMWF_FWI_",
                                             gsub("-", "",
                                                  as.character(day10)),
                                             "_1200_hr_fwi.nc")))[[10]]
    dfx$HRES_d10 <- raster::extract(x = HRES10, y = spdf)
    rm(HRES2, HRES6, HRES10)

    # Remove NAs
    dfx <- dfx %>% filter(complete.cases(.))
    # Re-define spatial point
    spdf <- dfx[, c("id", "lat", "long")]
    # Promote data.frame to spatial
    coordinates(spdf) = ~long + lat

    # Get data from ENS forecasts
    nstations <- dim(spdf@data)[1]
    mat2 <- matrix(NA, nrow = nstations, ncol = 51)
    mat6 <- matrix(NA, nrow = nstations, ncol = 51)
    mat10 <- matrix(NA, nrow = nstations, ncol = 51)
    for (j in 1:51){

      # Extract the modelled FWI from ENS
      ens_member <- sprintf("%02d", j - 1)
      # Day2
      ENS2 <- raster::brick(file.path(forecasts_folder,
                                       paste0("ECMWF_FWI_",
                                              gsub("-", "",
                                                   as.character(day2)),
                                              "_1200_", ens_member,
                                              "_fwi.nc")))[[2]]
      mat2[, j] <- raster::extract(x = ENS2, y = spdf)
      # Day6
      ENS6 <- raster::brick(file.path(forecasts_folder,
                                       paste0("ECMWF_FWI_",
                                              gsub("-", "",
                                                   as.character(day6)),
                                              "_1200_", ens_member,
                                              "_fwi.nc")))[[6]]
      mat6[, j] <- raster::extract(x = ENS6, y = spdf)
      # Day10
      ENS10 <- raster::brick(file.path(forecasts_folder,
                                       paste0("ECMWF_FWI_",
                                              gsub("-", "",
                                                   as.character(day10)),
                                              "_1200_", ens_member,
                                              "_fwi.nc")))[[10]]
      mat10[, j] <- raster::extract(x = ENS10, y = spdf)
    }
    # Populate CRPS for Day 2
    dfx$CRPS_d2 <- veriApply(verifun = "EnsCrps", fcst = mat2, obs = dfx$OBS)
    # Populate CRPS for Day 6
    dfx$CRPS_d6 <- veriApply(verifun = "EnsCrps", fcst = mat6, obs = dfx$OBS)
    # Populate CRPS for Day 10
    dfx$CRPS_d10 <- veriApply(verifun = "EnsCrps", fcst = mat10, obs = dfx$OBS)

    # Append to table
    df_used <- dplyr::bind_rows(df_used, dfx)

    saveRDS(df_used, "../data/df_geff_era5_hres_crps.rds")
  }
}

df_used <- readRDS("../data/df_geff_era5_hres_crps.rds")
```

```{r Figure3, echo = FALSE, fig.width = 13, fig.height = 6, out.width = "100%", fig.cap = "\\label{fig:Figure3}Comparison between modelled FWI and observed FWI value. FWI are calculated from using ECMWF reanalysis (ERA5) and forecasts at different lead times. The box plots are used to describe the distribution of values accross the observation points for the 2017 Mean Bias (left panel) and the 2017 mean absolute error (right panel)."}
d <- df_used %>%
  select(OBS, ERA5, HRES_d2, HRES_d6, HRES_d10) %>%
  filter(complete.cases(.)) %>% # Remove NAs
  reshape2::melt(id.vars = "OBS") %>%
  filter(OBS < 250, value < 250) %>%
  mutate(BIAS = OBS - value,
         MAE = abs(OBS - value)) %>%
  rename(modelled_type = variable, modelled_value = value) %>%
  reshape2::melt(id.vars = c("fwiobs", "modelled_type", "modelled_value"))

ggplot(d, aes(x = modelled_type, y = value, fill = modelled_type)) +
  facet_wrap( ~ variable, ncol = 2) +
  geom_boxplot(color="black", size=0.2,  outlier.shape = NA) +
  theme_bw() + ylim(-2, 12) +
  xlab("") + ylab("") +
  scale_fill_discrete_qualitative() +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c("ERA5", "HRES_D2", "HRES_D6", "HRES_D10"))
```

Despite its importance  the analysis performed using the synop network is still pointwise and does not cover all the regions where fires are relevant. Moreover mean biases and MAE are based on the high resolution forecast. These skill metrics do not provide information about the performance of the ensemble forecasting as a whole.

```{r Figure4, echo = FALSE, fig.width = 13, fig.height = 6, out.width = "100%", fig.cap = "\\label{fig:Figure4}PUT here crps skills scores."}
d <- df_used %>%
  select(region, CRPS_d2, CRPS_d6, CRPS_d10) %>%
  rename(Region = region, Day2 = CRPS_d2, Day6 = CRPS_d6, Day10 = CRPS_d10) %>%
  reshape2::melt(id.vars = "Region") %>%
  filter(complete.cases(.)) %>%
  rename(CRPS = value) %>%
  group_by(Region, variable) %>%
  summarise(CRPS = mean(CRPS))

ggplot(d, aes(x = variable, y = CRPS, colour = Region)) +
  geom_point() +
  xlab("") +
  scale_fill_manual(name="Region")
```

\subsection{Skill in detecting fire events}
